Deciphering enhancer rules underlying early human neuronal and neural crest development.

### 2025-03-24 manuscript.tex
More on discussion.

\subsection* {
  Four to five distinct transcription factor binding sites in a region of around 240 base pairs is sufficient to encode
  all cell states related to neuronal progenitors, early neuronal differentiation and neural crest development.
}

@@

% enhancers

% human tfs paper: https://www.sciencedirect.com/science/article/pii/S0092867418301065?via%3Dihub

% https://www.cell.com/fulltext/s0092-8674(12)01496-1: also about dimerization and that motif "flanks" are important for
% different TFs
% also: https://www.nature.com/articles/s41588-022-01048-5 about flank

%% Enhancers are clusters of TFBS
%%%% https://www.cell.com/cell/fulltext/S0092-8674(16)31251-X
%%%% https://www.cell.com/molecular-cell/fulltext/S1097-2765(22)01215-1
%%%% https://www.sciencedirect.com/science/article/pii/S0959440X21000993

%% Specificity through dimerization
%%%% https://www.cell.com/cell/fulltext/S0092-8674(23)01438-1
%%%% https://pmc.ncbi.nlm.nih.gov/articles/PMC6961570/

@@

%% pioneer
% https://www.nature.com/articles/s41588-018-0140-x
% https://elifesciences.org/articles/73358
% https://www.cell.com/cell/fulltext/S0092-8674(15)00304-9
% https://www.biorxiv.org/content/10.1101/2024.05.24.595749v3.abstract
% https://www.nature.com/articles/s41576-023-00648-z
% https://www.nature.com/articles/s41586-020-2195-y

%% Low affinity/suboptimization:
%%%% https://www.nature.com/articles/s41586-023-06922-8
%%%% https://www.science.org/doi/10.1126/science.aac6948
%%%% https://pmc.ncbi.nlm.nih.gov/articles/PMC11160683/
%%%% https://www.sciencedirect.com/science/article/abs/pii/S0070215315001842

@@

% https://www.pnas.org/doi/epub/10.1073/pnas.2411887121
% https://www.nature.com/articles/s41594-024-01235-4

@@

>>>>

\subsection* {
  Four to five distinct transcription factor binding sites in a region of around 140 base pairs is sufficient to encode
  all cell states related to neuronal progenitors, early neuronal differentiation and neural crest development.
}

@@

The enhancer code remains complicated to study\cite{wassermanAppliedBioinformaticsIdentification2004} especially
compared to the genetic code\cite{kimDecipheringMultiscaleQuantitative2023, deboerHoldOutGenome2024}. Nonetheless, there
are some recurrent patterns that are shared among enhancers of different cell types. In particular, enhancers underlying
DV patterning, early neuronal fate acquisition and neural crest tend to consist out of clusters of 4-5 different
TFBS\cite{kimDecipheringMultiscaleQuantitative2023, longEverChangingLandscapesTranscriptional2016} that are located
within one nucleosome depleted regions (+/-147bp). That is to say, cooperative binding of multiple TFs seem to underlie
the cell type specificity of enhancers\cite{lambertHumanTranscriptionFactors2018}. Here, we see most evidence for a
model of indirect cooperativity\cite{longEverChangingLandscapesTranscriptional2016} (i.e., flexible billboard enhancer
model\cite{arnostiTranscriptionalEnhancersIntelligent2005}). However, for multiple cell types we also observed evidence
of direct coooperativity between TFs in the form of protein-protein interactions between two TFs. This was the case for
TWIST1 and ALX4 heterodimer in facial mesenchyme\cite{kimDNAguidedTranscriptionFactor2024} and GRHL1/2 dimers, TFAP2A and TFAP2B/C  heterodimer\cite{rothsteinHeterodimerizationTFAP2Pioneer2020} and SOX10 homodimers in neural crest. For all these TFs we found dimer motifs with a fixed number nucleotides in between the two TFBS instances. Of note, for SOX10 we observed three different dimer motif with either 3, 4 or 5 bps spacing with those instances with 4 bps spacing being the most accessible (SUPPLEMENT FIG). Finally, we also observed evidence of pioneer activity, where binding of one TF can facilitate the subsequent binding of other TFs to the same enhancer\cite{longEverChangingLandscapesTranscriptional2016}. This was the case for FOXA2 in the floorplate, consistent with previous findings of FOXA TFs\cite{iwafuchi-doiPioneerTranscriptionFactor2016, iwafuchiGeneNetworkTransitions2020}. Others have hypothesized, however, that pioneer activity is not a qualitative feature of a TF but rather dependent on the affinity of interactions between TF and DNA\cite{hansenTestPioneerFactor2022}. Similarly, we observed that FOXA2 binding sites that are specific to the floorplate contain either less FOXA2 binding sites and/or binding sites of lower affinity. The use of suboptimal TFBS is a reoccurring feature of cell type specific enhancers\cite{farleySuboptimizationDevelopmentalEnhancers2015, crockerChapterTwentySevenSoft2016, limAffinityoptimizingEnhancerVariants2024, naqviTransferLearningReveals2025}.

Our results indicate that enhancers underlying neuronal progenitor AP-patterning form an exception to the rule that enhancer
consist out of clusters of 4-5 different TFBS. Homeotypic clusters of 3-4 homeobox instances were sufficient to both create and identify enhancers specific to a certain AP domain. Similar findings were already reported in fly\cite{crockerLowAffinityBinding2015}. It has been a long standing question how homeodomain TFs achieve specific binding\cite{mannChapter3Hox2009, hubertHoxGenesDevelopment2023} given that all homeodomain TFs seem to bind the same \textit{TAAT} motif\cite{bergerVariationHomeodomainDNA2008, noyesAnalysisHomeodomainSpecificities2008}. It has been hypothesized that specificity is mainly achieved through cooperative binding with co-factors\cite{mannChapter3Hox2009}. Indeed, we observe this for EN1 and HOXB3 that are predicted to co-bind with PAX8 and X respectively forming specific dimer motifs. In contrast, we suggest a novel mode where specificity is achieved through the homeodomain itself for TFX and TFY that preferentially bind \textit{TAATTA} and \textit{TAATCC} sites. Thus, additional nucleotides, next to the core \textit{TAAT}, form a TFBS that is both necessary and sufficient for genome wide cell type specificity. In general, although there are only around 100 different DNA binding domains across human TFs\cite{lambertHumanTranscriptionFactors2018} and thus many of the 1,639 known human TFs recognize similar motifs\cite{lambertHumanTranscriptionFactors2018} there are slight variations in preferential TFBS of TFs of the same family\cite{lambertHumanTranscriptionFactors2018, jolmaDNABindingSpecificitiesHuman2013, dealmeidaDeepSTARRPredictsEnhancer2022} that can be important for cell type specific binding.

@@

%% position of the binding sites vis a vi nucleosomes
% https://www.pnas.org/doi/abs/10.1073/pnas.1804663115

@@

@@

% co-factors (limitation)
% "new" studies
% https://www.biorxiv.org/content/10.1101/2022.08.26.505496v1.abstract
% https://www.cell.com/cell/fulltext/S0092-8674(20)31541-5?uuid=uuid%3A51e14242-dce3-432c-8521-14d9a94b0381
% https://www.cell.com/molecular-cell/fulltext/S1097-2765(21)01070-4?uuid=uuid%3Afbec3fc1-2707-4902-bda8-f196da451272
% https://www.science.org/doi/full/10.1126/science.adf6149
% focus on single enhancers and not entire locus
%
% from lambert et al
%%TFs have traditionally been classified as either ‘‘activators’’ or ‘‘repressors’’; however, this notion has been repeatedly questioned. Many TFs can recruit multiple cofactors that have opposite effects (Frietze and Farnham, 2011; Rosenfeld et al., 2006;
% Schmitges et al., 2016), dependent on the local sequence
%context and availability of cofactors (Meijsing et al., 2009;
%Wong and Struhl, 2011).
